Decomposing biodiversity change to processes of extinction, colonization, and persistence across scales

### 2023-09-26 scripts/ms_scaling_changes.rmd
revision

```{r}
## Add the 100x100 km
load(file = "../data/data_to_predict/atlas_data_topredict_with10000sqkm.Rdata")
## Prepare the dataset
pred10000sqkm <-
  atlas_data_topredict_with10000sqkm %>%
  filter(spatial_scale == 10000) %>% 
  rbind(atlas_data_topredict_with10000sqkm %>%
          filter(spatial_scale == 10000) %>% 
          mutate(Date = 2002)) %>% 
    rbind(atlas_data_topredict_with10000sqkm %>%
          filter(spatial_scale == 10000) %>% 
          mutate(Date = 2015)) %>% 
  dplyr::select(-c(spatial_scale))
## Make the prediction
pred10000sqkm$pred_sr <- round(predict(models_list_atlas$xgboost$finalModel, newdata = as.matrix(pred10000sqkm)))
# pred10000sqkm$KVADRAT <- atlas_data_topredict$KVADRAT[atlas_data_topredict$spatial_scale == 10000]
## And compute the trend
## M2 --> M4 
slope_10000_M2M4 <-
pred10000sqkm %>% 
  group_by(elevation) %>% 
  do({
    mod <-  lm(pred_sr ~ Date, data = .)
    data.frame(slope = coef(mod)[2])
  }) %>% 
  filter(!is.na(elevation))
## M2 --> M3
slope_10000_M2M3 <-
pred10000sqkm %>% 
  filter(Date != 2015) %>% 
  group_by(elevation) %>% 
  do({
    mod <-  lm(pred_sr ~ Date, data = .)
    data.frame(slope = coef(mod)[2])
  }) %>% 
  filter(!is.na(elevation))
## M3 --> M4
slope_2150_M3M4 <-
pred2150sqkm %>% 
  filter(Date != 1987) %>% 
  group_by(elevation) %>% 
  do({
    mod <-  lm(pred_sr ~ Date, data = .)
    data.frame(slope = coef(mod)[2])
  }) %>% 
  filter(!is.na(elevation))

@@

```{r, warning=FALSE, message=FALSE}
# M2M4 ####
pdf("../figures/sr_spatialscaling_M2M4.pdf",
    height = 5, width = 5.5)
# Data management
slope_JPSP_M2M4 %>% 
  dplyr::rename(area = AREA) %>% 
  dplyr::select(slope, area) %>% 
  mutate(area = ifelse(area < 0.2, 0.13, 2.51)) %>% 
  rbind(
    slope_130_M2M4 %>% 
      dplyr::select(slope) %>% 
      mutate(area = 130) %>% 
      rbind(
        slope_530_M2M4 %>% 
          dplyr::select(slope) %>% 
          mutate(area = 530)
      ) %>% 
      rbind(
        slope_2150_M2M4 %>% 
          dplyr::select(slope) %>% 
          mutate(area = 2150)
      ) %>% 
      # rbind(
      #   slope_2150_M2M4 %>% 
      #     dplyr::select(slope) %>% 
      #     mutate(area = 10000)
      # ) %>% 
      rbind(
        slope_CR_M2M4 %>% 
          dplyr::select(slope) %>% 
          mutate(area = 83900)
      )
  ) %>% 
  ## Plot
  ggplot() +
  geom_boxplot(aes(area, slope, group = area))+
  # geom_boxplot(aes(area, slope, group = area), data = . %>% filter(area == 10000), 
  #              color = "blue", width = 0.3, size = 1)+
  scale_x_log10()+
  xlab("Spatial grain (km²)")+
  ylab("Species Richness change\nper year")+
  ylim(-2,2)+
    theme(panel.background = element_blank(),
        panel.grid.major = element_line(colour = "lightgrey"),
        axis.text.x = element_markdown())+
  stat_summary(aes(area, slope), fun.y = median, geom = 'line', linetype = 2, size = 1,  color = "blue")+
  stat_summary(aes(area, slope), fun = mean,
               geom = "point", col = "red", size = 1)+
  stat_summary(aes(area, slope), fun.data = mean_se,
               geom = "errorbar", size = .5, width = .1, color = "red")+
  geom_hline(yintercept = 0)+
  annotation_logticks(sides = "b")+
  coord_cartesian(clip = "off")+
  theme(text = element_text(size = 16))
dev.off()

@@

```{r}
pdf("../figures/pCEP_2periods.pdf",
    height = 5, width = 5.5)
for(THE_metric in c("col", "ext")){
  print(predicted_pCEP_BBS %>% 
  mutate(spatial_scale = ifelse(AREA < 0.2, 0.13, 2.51)) %>%
  mutate(period = fct_relevel(period, c("M2M4", "M2M3", "M3M4"))) %>% 
  dplyr::select(spatial_scale, pred, period, metric) %>% 
  rbind(predicted_pCEP_atlas %>% dplyr::select(spatial_scale, pred, period, metric)) %>% 
  filter(pred >= 0 & pred <= 1) %>% 
  filter(period != "M2M4") %>%
  filter(metric == THE_metric) %>%
  ggplot()+
  geom_boxplot(aes(spatial_scale, pred, group = interaction(spatial_scale, period), color = period))+
  stat_summary(aes(spatial_scale, pred, color = period, linetype = period), fun.y = median, geom = 'line', linetype = 2, size = 1)+
  scale_x_log10()+
  annotation_logticks(sides = "b")+
  xlab("Spatial grain (km²)")+
  ylab(THE_metric)+
  ylim(0, .5)+
  labs(color = "Period", shape = "Period", linetype = "Period")+
  theme_bw()+
  theme(text = element_text(size = 16))+
  theme(panel.background = element_blank(),
        legend.position = "none",
        panel.grid.major = element_line(colour = "lightgrey"),
        axis.text.x = element_markdown()))
}

## Variable importance (Appendix B)

```{r}
CEP_atlas_cart_data <- readRDS("../data/data_model_fitting/CEP_atlas_cart_data.rds")
CEP_atlas_cart_data$metric[CEP_atlas_cart_data$metric == "proba_colonized"] <- "proba_col"

CEP_JPSP_cart_data <- readRDS("../data/data_model_fitting/CEP_JPSP_cart_data.rds")
CEP_JPSP_cart_data <- CEP_JPSP_cart_data %>% 
  filter(metric != "proba colonization") %>% 
  mutate(metric = case_when(
    metric == "colonization" ~ "col",
    metric == "extinction" ~ "ext",
    metric == "persistence" ~ "pers",
    metric == "proba persistence" ~ "proba_pers",
    metric == "proba extinction" ~ "proba_ext",
    metric == "proba colonized" ~ "proba_col",
  ))

pdf("../figures/correlations.pdf",
    height = 8.27, width = 11.69)
## For Atlas model
plot(atlas_cart_data[,names(atlas_cart_data) != "sr"],
     main = "Correlation - CzAtlas - species richness predictors")

ggplot(data = varImp(models_list_atlas$xgboost), aes(x = reorder(rownames(importance), Overall), y = Overall)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(x = "Predictor Variables", y = "Importance") +
  ggtitle("Variable Importance Plot - CzAtlas - XGBoost algorithm") +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

CEP_atlas_cart_data %>% filter(metric == "ext") %>% 
    select(-quadrat, -metric, -trials, -value) %>% 
    plot(main = "Correlation - CzAtlas - colonization/extinction/persistence")

for(THE_metric in unique(CEP_atlas_cart_data$metric)){
  if(!grepl("proba_", THE_metric)){
    print(
      ggplot(data = varImp(nCEP_atlas_models[[THE_metric]][["xgboost"]]), aes(x = reorder(rownames(importance), Overall), y = Overall)) +
      geom_bar(stat = "identity", fill = "blue") +
      labs(x = "Predictor Variables", y = "Importance") +
      ggtitle(paste0("Variable Importance Plot - CzAtlas - ", THE_metric)) +
      theme_bw()+
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    )
  }else{
    print(
      ggplot(data = varImp(pCEP_atlas_models[[gsub("proba_","",THE_metric)]][["xgboost"]]), aes(x = reorder(rownames(importance), Overall), y = Overall)) +
      geom_bar(stat = "identity", fill = "blue") +
      labs(x = "Predictor Variables", y = "Importance") +
      ggtitle(paste0("Variable Importance Plot - CzAtlas - ", THE_metric)) +
      theme_bw()+
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
    )
  }
}

## For CzBBS model
plot(JPSP_cart_data[,names(JPSP_cart_data) != "sr"],
     main = "Correlation - CzBBS - species richness predictors")

ggplot(data = varImp(models_list_JPSP$rf), aes(x = reorder(rownames(importance), Overall), y = Overall)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(x = "Predictor Variables", y = "Importance") +
  ggtitle("Variable Importance Plot - CzBBS Data - random forest algorithm") +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

CEP_JPSP_cart_data %>% 
  filter(metric == "ext") %>%
  dplyr::select(-metric) %>% 
  filter(!is.na(value)) %>% 
  plot(main = paste0("Correlation - CzBBS - colonization/extinction/persistence predictors"))

for(THE_metric in unique(CEP_JPSP_cart_data$metric)){
  if(!grepl("proba_", THE_metric)){
    print(
      ggplot(data = varImp(nCEP_BBS_models[[THE_metric]][["rf"]]), aes(x = reorder(rownames(importance), Overall), y = Overall)) +
      geom_bar(stat = "identity", fill = "blue") +
      labs(x = "Predictor Variables", y = "Importance") +
      ggtitle(paste0("Variable Importance Plot - CzBBS - ", THE_metric)) +
      theme_bw()+
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    )
  }else{
    print(
      ggplot(data = varImp(pCEP_atlas_models[[gsub("proba_","",THE_metric)]][["xgboost"]]), aes(x = reorder(rownames(importance), Overall), y = Overall)) +
      geom_bar(stat = "identity", fill = "blue") +
      labs(x = "Predictor Variables", y = "Importance") +
      ggtitle(paste0("Variable Importance Plot - CzBBS - ", THE_metric)) +
      theme_bw()+
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
    )
  }
}

>>>>

```{r}
## Add the 100 x 100 Km
load(file = "../data/data_to_predict/atlas_data_topredict_with10000sqkm.Rdata")

## Prepare the dataset
pred10000sqkm <-
  atlas_data_topredict_with10000sqkm %>%
  filter(spatial_scale == 10000) %>%
  rbind(atlas_data_topredict_with10000sqkm %>%
  filter(spatial_scale == 10000) %>%
  mutate(Date = 2002)) %>%
  rbind(atlas_data_topredict_with10000sqkm %>%
  filter(spatial_scale == 10000) %>%
  mutate(Date = 2015)) %>%
  dplyr::select(-c(spatial_scale))
## Make the prediction
pred10000sqkm$pred_sr <- round(predict(models_list_atlas$xgboost$finalModel, newdata = as.matrix(pred10000sqkm)))
# pred10000sqkm$KVADRAT <- atlas_data_topredict$KVADRAT[atlas_data_topredict$spatial_scale == 10000]
## And compute the trend
## M2 --> M4
slope_10000_M2M4 <-
  pred10000sqkm %>%
  group_by(elevation) %>%
  do({
  mod <-  lm(pred_sr ~ Date, data = .)
  data.frame(slope = coef(mod)[2])
}) %>%
  filter(!is.na(elevation))

@@

```{r, warning=FALSE, message=FALSE}
# M2M4 ####
pdf("../figures/sr_spatialscaling_M2M4.pdf",
    height = 5, width = 5.5)
# Data management
slope_JPSP_M2M4 %>% 
  dplyr::rename(area = AREA) %>% 
  dplyr::select(slope, area) %>% 
  mutate(area = ifelse(area < 0.2, 0.13, 2.51)) %>% 
  rbind(
    slope_130_M2M4 %>% 
      dplyr::select(slope) %>% 
      mutate(area = 130) %>% 
      rbind(
        slope_530_M2M4 %>% 
          dplyr::select(slope) %>% 
          mutate(area = 530)
      ) %>% 
      rbind(
        slope_2150_M2M4 %>% 
          dplyr::select(slope) %>% 
          mutate(area = 2150)
      ) %>% 
      rbind(
        slope_10000_M2M4 %>%
          dplyr::select(slope) %>%
          mutate(area = 10000)
      ) %>% 
      rbind(
        slope_CR_M2M4 %>% 
          dplyr::select(slope) %>% 
          mutate(area = 83900)
      )
  ) %>% 
  ## Plot
  ggplot() +
  geom_boxplot(aes(area, slope, group = area))+
  geom_boxplot(aes(area, slope, group = area), data = . %>% filter(area == 10000),
               color = "blue", width = 0.3, size = 1)+
  scale_x_log10()+
  xlab("Spatial grain (km²)")+
  ylab("Species Richness change\nper year")+
  ylim(-2,2)+
    theme(panel.background = element_blank(),
        panel.grid.major = element_line(colour = "lightgrey"),
        axis.text.x = element_markdown())+
  stat_summary(aes(area, slope), fun.y = median, geom = 'line', linetype = 2, size = 1,  color = "blue")+
  stat_summary(aes(area, slope), fun = mean,
               geom = "point", col = "red", size = 1)+
  stat_summary(aes(area, slope), fun.data = mean_se,
               geom = "errorbar", size = .5, width = .1, color = "red")+
  geom_hline(yintercept = 0)+
  annotation_logticks(sides = "b")+
  coord_cartesian(clip = "off")+
  theme(text = element_text(size = 16))
dev.off()

@@

```{r}
pdf("../figures/pCEP_2periods.pdf",
    height = 5, width = 5.5)
for(THE_metric in c("col", "ext")){
  print(predicted_pCEP_BBS %>% 
  mutate(spatial_scale = ifelse(AREA < 0.2, 0.13, 2.51)) %>%
  mutate(period = fct_relevel(period, c("M2M4", "M2M3", "M3M4"))) %>% 
  dplyr::select(spatial_scale, pred, period, metric) %>% 
  rbind(predicted_pCEP_atlas %>% dplyr::select(spatial_scale, pred, period, metric)) %>% 
  filter(pred >= 0 & pred <= 1) %>% 
  filter(period != "M2M4") %>%
  filter(metric == THE_metric) %>%
  ggplot()+
  geom_boxplot(aes(spatial_scale, pred, group = interaction(spatial_scale, period), color = period))+
  stat_summary(aes(spatial_scale, pred, color = period, linetype = period), fun.y = median, geom = 'line', linetype = 2, size = 1)+
  scale_x_log10()+
  annotation_logticks(sides = "b")+
  xlab("Spatial grain (km²)")+
  ylab(THE_metric)+
  ylim(0, .5)+
  labs(color = "Period", shape = "Period", linetype = "Period")+
  theme_bw()+
  theme(text = element_text(size = 16))+
  theme(panel.background = element_blank(),
        legend.position = "none",
        panel.grid.major = element_line(colour = "lightgrey"),
        axis.text.x = element_markdown()))
}
dev.off()
```

## Variable importance (revision)

```{r}
pdf("../figures/correlations.pdf",
    height = 5, width = 5.5)
plot(JPSP_cart_data[,names(JPSP_cart_data) != "sr"])
plot(atlas_cart_data[,names(atlas_cart_data) != "sr"])

## For Atlas XGBoost model
ggplot(data = varImp(models_list_atlas$xgboost), aes(x = reorder(rownames(importance), Overall), y = Overall)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(x = "Predictor Variables", y = "Importance") +
  ggtitle("Variable Importance Plot - CzAtlas - XGBoost algorithm") +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
## For CzBBS XGBoost model
ggplot(data = varImp(models_list_JPSP$xgboost), aes(x = reorder(rownames(importance), Overall), y = Overall)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(x = "Predictor Variables", y = "Importance") +
  ggtitle("Variable Importance Plot - CzBBS Data - XGBoost algorithm") +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
